FraCaS: Temporal Analysis

### 2020-06-23 paper2020/paper.tex
revise sem. sec

In montagovian semantics, (intransitive) verbs are one-place
predicates; in types, they are functions from entities to propositions
($e \to t$).  Our basic approach is to generalise the interpretation of 
verbs, so that it takes two
additional time parameters, one corresponding with the starting time
of the action and one corresponding to its stopping time ($(e × time × time) \to t$). For example,
if John walked between $t_0$ and $t_1$, we would have:
$walk(john,t_0,t_1)$. From now on we will call an interval of time
points $[t_0,t_1]$ a timespan. Every timespan $[t_0,t_1]$ has the
property $t_0 ≤ t_1$ (it starts no later than it stops). (We are thus
using a simple newtonian model of time, corresponding to a layman
intuition of a linear constant flow of time.)

In principle, common nouns and adjectives should undergo the same
procedure. For simplicity we will however only consider verbs from now
on. (In fact in our implementation we chose not to modify nouns nor
adjectives to limit the increase in complexity of the formulas.)

@@

This temporal context is an \emph{optional} timespan. That is, it can
be an timespan or an explicitly unspecified context.
\newcommand\nospan{\ensuremath{-}}
\newcommand\anyspan{\ensuremath{\mathbf{?}}}\todo{better notation} The timespan in the context is
optional because, in certain situations, the semantics is different
depending on whether a timespan has been specified externally or not,
as we explain below. A non-present timespan will be represented as
\nospan{}. If a semantic function does not depend on the temporal
context at all, we will write \anyspan{} instead.

@@

\newcommand\varid[1]{\mathnormal{#1}}
%
The principal non-trivial manipulator of timespans are tense
markers. In our syntax, inherited from GF, tenses are represented
syntactically as an attribute of the clause. An illustration of a
past-tense clause and its interpretation follows. Notice in particular
the $past$ argument to $useCl$.

\onelingex{A scandinavian won the nobel prize.\label{ex:nobel}
\\
{\small\begin{multline*}
\varid{useCl}\, \varid{past}\, \varid{pPos}\\ (\varid{predVP}\, (\varid{detCN}\, (\varid{detQuant}\, \varid{indefArt}\, \varid{numSg})\\ \varid{scandinavian\_CN})\\ (\varid{complSlash}\, (\varid{slashV2a}\, \varid{win\_V2})\\ (\varid{detCN}\, (\varid{detQuant}\, \varid{indefArt}\, \varid{numSg})\\ \varid{nobel\_prize\_CN})))
\end{multline*}
}}
In our semantics we deal only with present and past
tenses (simple and continuous). Indeed we find that FraCas
does not exercise additional specific tenses (if a more complicated
tense is used, the additional information is also carried by adverbs
or adverbial phrases, in a more specific way.) It is not difficult to see how other tenses can be captured under the same general framework.

Even though we will discuss a refinement to handle the past continuous
at the end of this section, the procedure to handle tense annotations
is as follows:
\begin{itemize}
\item If the tense is the past, and the temporal context is
  unspecified, then we locally quantify over a time interval
  $[t_0,t_1]$, such that $t_1 < now$, where $now$ is a logical
  constant representing the current timepoint. The temporal context
  then becomes this interval.
\item If the tense is the present and the temporal context is
  unspecified, then the temporal context becomes the simple
  $(now,now)$ interval.
\item If the temporal context is specified (for example due to the presence of an adverb or an
  averbial clause, such as ``before James swam''), then the tense does
  not create a new interval, but it may constrain it. Typically, a
  past tense adds the constraint that the temporal context ends before
  the timepoint $now$.
\end{itemize}

@@

\begin{enumerate}[align=left,font=\itshape]
\item[exact] For such adverbs, an exact interval is
  provided. Typical such adverbs specify in fact a single point in
  time (so the start and the end of the interval coincide).

@@

  \begin{multline*}
  ⟦\text{in 1992, \(s\)}⟧(\anyspan) = \\ ∃t_1,t_2.  [t_1,t_2] ⊆ 1992, ⟦s⟧(t_1,t_2)
\end{multline*}
In the FraCaS test suite, we normally do not find several
time-modifying adverbs modifying a single verb. Indeed, sentences such
as ``in 1992, in 1991 john wrote a novel'' are infelicitous. This
justifies ignoring the input timespan in the above interpretation --
we're in particular not interested in modeling felicity with our
semantics, only giving an accurate semantics when the input is
felicitous.
\item[universally quantifying] A few adverbs introduce intervals via a
  universal quantification (sometimes with a constraint). Examples
  include ``always'' and ``never''.

@@

To be able to conclude that there is entailment, as the testsuite expects, we have to make sure
that the two occurences of ``Jones left'' (in \textbf{P1} and \textbf{P2}) refer to the
same time intervals.

For this purpose we postulate \emph{unicity of action} for certain time-dependent
propositions:

\begin{multline*}
  \constant{unicity}_P : P (t_1,t_2) → P (t_3,t_4) → \\ (t_1 = t_3) ∧ (t_2 = t_4)
  \label{eq:unicity}
\end{multline*}

@@

(The difference between actitivty and accomplishments on the one and
achievement on the other hand is that for the latter, time intervals
can be assumed to be of nil duration. In reality, this is an
oversimplification as achievements are usually of short duration, but
not nil. However, this appears to play little role, if any, in our
analysis, and given the its task-driven nature as well.)

Unicity of action appears to be a non-logical principle. Indeed, it is
quite possible that ``Jones left'' several times. However, it seems
that this principle is never contradicted by the testsuite. As such,
even though unicity of action is only a pragmatic rule, it can be
taken as a valid one \emph{by default}: it is only when we have a
sufficiently constrained situation that one should reject it. Consider
the following discourse:
\begin{enumerate}[parsep=0pt,itemsep=1pt,label=(\arabic*)]
\item Smith left at 1pm.
\item Smith went to his appointment with the lawyer.
\item Smith left at 4pm.
\end{enumerate}
One would normally not say that there is contradiction. However if the
middle sentence were not present, a contradiction should be
flagged. We leave such discourse analysis as future work, and simply
apply unicity of action everywhere.

@@

Indeed, from \textbf{P3} we get that Smith was in paris between May
5th and May 15th. Because "being in Paris" is stative, we also get that
Smith was in Paris in any subinterval. Contrary to unicity of action,
subsumption is always valid.

@@

We treat verbs in the progressive form as different semantically. For
example, ``John was writing a book'' is encoded as
$∃(t_1, t_2). t_1≤ t_2, t_2 ≤ now, PROG\_write(John,book,t_1,t_2)$,
while ``John wrote a book'' is encoded as
$∃(t_1, t_2). t_1≤ t_2, t_2 ≤ now, write(John,book,t_1,t_2)$. This is
necessary, because $PROG\_write$ is subject to
$\constant{subsumption}$: if John is writing in interval $[t_1,t_2]$
then he is writing in any subinterval. However, the same cannot be
said of $write$: the end-points of the interval indicates the time
needed to complete the achievement. (For example, ``John wrote a book
in 1993'' neither entails ``John wrote a book in January 1993'' nor
``John wrote a book in December 1993''.) That is, $write$ is on the
contrary subject to $\constant{unicity}$. Finally, we also have
$write(x,y,t_1,t_2) → PROG\_write(x,y,t_1,t_2)$. That is, the
achievement variant implies the stative variant, in the same interval.
Consequently we get the entailment from ``John wrote a book'' to
``John was writing a book''.

@@

\section{Conclusions and Future Work}
\label{sec:conclusion}
We have presented a first attempt for a computational approach dealing
with the temporal section of the FraCaS test suite. This is the first
time a system is ran against this section, at the same time, this
concludes our work in handling the whole of the FraCaS test suite
using a combination of a number of well-studied tools: type theory,
parsing using the Grammatical Framework (GF), Monadic Dynamic
Semantics and proof assistant technology (Coq). It achieves an
accuracy of 0.71 on the Temporal Section and 0.78 overall. One of the
things to be looked at is fixing the issues associated with parts of
the test suite that ``broke'' so to say when the temporal analysis was
introduced. Some of these have been already mentioned: interaction of
the temporal variables with group readings and negation.

>>>>

In montagovian semantics, (intransitive) verbs are one-place
predicates; in types, they are functions from entities to propositions
($e \to t$).  Our basic approach is to generalise the interpretation of 
verbs, so that it takes two
additional time parameters, one corresponding to the starting time
of the action and one corresponding to its stopping time ($(e × time × time) \to t$). For example,
if John walked between $t_0$ and $t_1$, we would have:
$walk(john,t_0,t_1)$. From now on we will call an interval of time
points $[t_0,t_1]$ a timespan. Every timespan $[t_0,t_1]$ has the
property $t_0 ≤ t_1$ (it starts no later than it stops). (We are thus
using a simple newtonian model of time, corresponding to a layman
intuition of a linear constant flow of time.)

In principle, common nouns and adjectives should undergo the same
procedure. For simplicity we will however only consider verbs from now
on. (In fact, even in our implementation we chose not to extend nouns
nor adjectives with timespan parameters. This choice limits the
increase in complexity of the formulas compared to non-temporal
semantics, at the expense of inaccuracy for a couple of problems in
the FraCaS test suite: problems 271 and 272 use a an adjective as a
copula which is subject to temporal reasoning.)

@@

This temporal context is an \emph{optional} timespan. That is, it can
be a timespan or an explicitly unspecified context.
\newcommand\nospan{\ensuremath{-}}
\newcommand\anyspan{\ensuremath{\mathbf{?}}}\todo{better notation} The timespan in the context is
optional because, in certain situations, the semantics is different
depending on whether a timespan has been specified externally or not,
as we explain below. A non-present timespan will be represented as
\nospan{}. If a semantic function does not depend on the temporal
context at all, we will write \anyspan{} instead.

@@

\newcommand\varid[1]{\mathnormal{#1}}
%
The principal non-trivial manipulators of timespans are tense
markers. In our syntax, inherited from GF, tenses are represented
syntactically as an attribute of clauses. An illustration of a
past-tense clause and its interpretation follows in
\cref{ex:nobel}. Notice in particular the $\varid{past}$ argument to
the $\varid{useCl}$ constructor.

\onelingex{A scandinavian won the nobel prize.\label{ex:nobel}
\\
{\small\begin{multline*}
\varid{useCl}\, \varid{past}\, \varid{pPos}\\ (\varid{predVP}\, (\varid{detCN}\, (\varid{detQuant}\, \varid{indefArt}\, \varid{numSg})\\ \varid{scandinavian\_CN})\\ (\varid{complSlash}\, (\varid{slashV2a}\, \varid{win\_V2})\\ (\varid{detCN}\, (\varid{detQuant}\, \varid{indefArt}\, \varid{numSg})\\ \varid{nobel\_prize\_CN})))
\end{multline*}
}}
%
In our semantics we deal only with present and past tenses (simple and
continuous). Indeed we find that FraCas does not exercise additional
specific tenses. (When a more complicated tense is used, the
additional information is also carried by adverbs or adverbial
phrases, in a more specific way). While we believe that many other
tenses can be captured under the same general framework, we leave a
detailed study to further work.

Even though we discuss a refinement to handle the past continuous
at the end of this section, the procedure to handle tense annotations
is as follows:
\begin{itemize}
\item If the tense is the past, and the temporal context is
  unspecified, then we locally quantify over a time interval
  $[t_0,t_1]$, such that $t_1 < now$, where $now$ is a logical
  constant representing the current timepoint. The temporal context
  then becomes this interval.
\item If the tense is the present and the temporal context is
  unspecified, then the temporal context becomes the simple
  $(now,now)$ interval.
\item If the temporal context is specified (for example due to the presence of an adverb or an
  averbial clause, such as ``before James swam''), then the tense does
  not create a new interval, but it may constrain it. Typically, a
  past tense adds the constraint that the temporal context ends before
  the timepoint $now$.
\end{itemize}

@@

\begin{enumerate}[align=left,font=\itshape]
\item[exact] For such adverbs, an exact interval is
  provided. In fact, such adverbs typically specify a single point in
  time (so the start and the end of the interval coincide).

@@

  \begin{multline*}
  ⟦\text{in 1992, \(s\)}⟧(\anyspan) = \\ ∃t_1,t_2.  [t_1,t_2] ⊆ 1992, ⟦s⟧(t_1,t_2)
\end{multline*}
In the FraCaS test suite, we normally do not find several
time-modifying adverbs modifying a single verb phrase. Indeed, sentences such
as ``in 1992, in 1991 john wrote a novel'' are infelicitous. This
justifies ignoring the input timespan in the above interpretation --
we are in particular not interested in modeling felicity with our
semantics, only giving an accurate semantics when the input is
felicitous.
\item[universally quantifying] A few adverbs introduce intervals via a
  universal quantification (sometimes with a constraint). Examples
  include ``always'' and ``never''.

@@

To be able to conclude that there is entailment, as the testsuite expects, we have to make sure
that the two occurences of ``Jones left'' (in \textbf{P1} and \textbf{P2}) refer to the
same time intervals.
%
For this purpose we postulate \emph{unicity of action} for certain time-dependent
propositions:

\begin{multline}
  \constant{unicity}_P : P (t_1,t_2) → P (t_3,t_4) → \\ (t_1 = t_3) ∧ (t_2 = t_4)
  \label{eq:unicity}
\end{multline}

@@

(The difference between actitivty and accomplishments on the one hand and
achievement on the other hand is that for the latter, time intervals
can be assumed to be of nil duration. In reality, this is an
oversimplification as achievements are usually of short duration, but
not nil. However, this plays little role in our
analysis. As far as we can tell the FraCaS test suite does exercise temporal semantics to
such a level of precision.)

Unicity of action appears to be a non-logical principle. Indeed, it is
quite possible that ``Jones left'' several times. However, it seems
that this principle is never contradicted by the testsuite. As such,
even though unicity of action is only a pragmatic rule, it can be
taken as a valid one \emph{by default}: it is only when we have a
sufficiently constrained situation that one should reject it. Consider
the following discourse:
\begin{enumerate}[parsep=0pt,itemsep=1pt,label=(\arabic*)]
\item Smith left at 1pm.
\item Smith went to his appointment with the lawyer.
\item Smith left at 4pm.
\end{enumerate}
One would normally not say that there is contradiction. However if the
middle sentence were not present, a contradiction should be
flagged. We leave such discourse analysis as future work, and simply
apply unicity of action everywhere: it is valid uniformly in the
FraCaS test suite for activity aspect classes.

@@

Indeed, from \textbf{P3} we get that Smith was in paris between May
5th and May 15th. Because ``being in Paris'' is stative, we also get that
Smith was in Paris in any subinterval. Contrary to unicity of action,
subsumption is always valid.

\paragraph{Class-modifying adverbs}
It should be noted that some adverbs can locally disable the
application of $\constant {subsumption}$. For example, problem 299
features the sentence ``Smith lived in Birmingham for exactly a
year''.  Even though ``live'' is normally stative, one can no longer
apply subsumption in the context of ``exactly a year'' --- this can be
done by propagating another context flag in the montegovian semantics
(in addition to the temporal context).

@@

We treat verbs in the progressive form as different semantically from
the non-progressive form. For example, ``John was writing a book'' is
encoded as
$∃(t_1, t_2). t_1≤ t_2, t_2 ≤ now, PROG\_write(John,book,t_1,t_2)$,
while ``John wrote a book'' is encoded as
$∃(t_1, t_2). t_1≤ t_2, t_2 ≤ now, write(John,book,t_1,t_2)$. This is
necessary, because $PROG\_write$ is subject to
$\constant{subsumption}$: if John is writing in the interval
$[t_1,t_2]$ then he is writing in any subinterval of
$[t_1,t_2]$. However, the same cannot be said of $write$: the
end-points of the interval indicate the time needed to complete the
achievement. (For example, ``John wrote a book in 1993'' neither
entails ``John wrote a book in January 1993'' nor ``John wrote a book
in December 1993''.) In fact, \emph{write}, in the non-progressive
from, is on the contrary subject to $\constant{unicity}$. Finally, we
also have $write(x,y,t_1,t_2) → PROG\_write(x,y,t_1,t_2)$. That is,
the achievement (or \emph{activity} in our terminology) variant implies the stative variant, in the
same interval.  Consequently we get the entailment from ``John wrote a
book'' to ``John was writing a book''.

@@

\section{Conclusions and Future Work}
\label{sec:conclusion}
\todo{summary of advances}

We have presented a first attempt for a computational approach dealing
with the temporal section of the FraCaS test suite. This is the first
time a system is ran against this section, at the same time, this
concludes our work in handling the whole of the FraCaS test suite
using a combination of a number of well-studied tools: type theory,
parsing using the Grammatical Framework (GF), Monadic Dynamic
Semantics and proof assistant technology (Coq). It achieves an
accuracy of 0.71 on the Temporal Section and 0.78 overall. One of the
things to be looked at is fixing the issues associated with parts of
the test suite that ``broke'' so to say when the temporal analysis was
introduced. Some of these have been already mentioned: interaction of
the temporal variables with group readings and negation.
